### 2024-03-27 thesis/appendix.tex
March 27th 2024

\begin{table}
    \caption{MAE Vehicle data}
    \begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    0 & 1.117714 & 0.606971 & 0.752197 & 0.594292 \\
    \bottomrule
    \end{tabular}
\end{table}

@@

\section{Kernel methods best practices}
Following, are reported a couple of considerations important to keep in mind when working with kernel methods.
\subsection{Data scaling}
\subsection{Data compression}
Nystrom, Cholesky

>>>>

\begin{table}
    \caption{MAE Vehicle data}
    \begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    0 & 1.117714 & 0.606971 & 0.752197 & 0.594292 \\
    \bottomrule
    \end{tabular}
\end{table}
What can be concluded from these numerical examples is that, on average kernel quantile regression yields better results than quantile forest \cite{meinshausen2006quantile} and gradient boosting machine quantile regression \cite {friedman2001greedy} in terms of the pinball loss as well as in terms of the mean absolute error.

@@

\section{Kernel methods best practices}
Following, are reported a couple of considerations important to keep in mind when working with kernel methods.

\subsection{Data normalization}
With data normalization we transform the range of the feautures to a standard scale.
Such preprocessing step is essential when employing distance based algorithms like svm or k-nearest neighbors. The rationale behind it is that by normalizing data we give an uniform weight to each feature in the learning process; in this way we do not favour larger scale features.
Examples of the most popular features scaling are:
\begin{itemize}
    \item Standard scaler= it computes the standard score z of a sample x
    \\
    z=$\frac{x-\mu}{\sigma}$.
    \item MinMax scaler= it maps every data sample to the range 0, 1.
    \\
    $\frac{x-\min(X)}{\max(X)-\min(X)}$
    \item Robust scaler= it scales features using statistics that are robust to outliers.
    Essentially, it subtracts the median and then scales the data according to the interquantile range.
\end{itemize}
Many other data scaling algorithms exists, we refer the reader to a thorough comparison \cite{scikitscalers}.
\\
We conclude this subsection with a custom example that motivates the need of feature scaling \cite{scikitscale_example}. The idea is to compare the results of modelling the data with k nearest neighbors on the unscaled data against the scaled data. The considered data is the wine recognition dataset \href{https://archive.ics.uci.edu/dataset/109/wine}{https://archive.ics.uci.edu/dataset/109/wine}. The goal for this dataset is recognising from whose cultivator the wine comes based on two features with a completely different scale. The first feature has values in the [0,1000] range while the second feature has values contained in [1,10].
The unscaled and scaled version are compared in figure \ref{fig:feature_scaler_example}
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/feature_scaler_example.png}
    \caption{importance of feature scaling}
    \label{fig:feature_scaler_example}
\end{figure}
What we can conclude from the image is that, the model trained on scaled data is greatly better than the other.
On the left, we can see that distances between categories are impacted solely by the larger scale feature. Conversely, on the right, we have that the two features contributing equally in determining the neighbors.

\subsection{Data compression}
When the number of points n is large, kernel methods suffer from high computational costs. 
Using kernel methods, we have that the storage cost is of the order of $O(n^2)$ while the computational cost for finding the solution is of the order $O(n^3)$.
A significant speed up can be obtained thanks to low rank approximation.

\subsubsection{Nystrom decomposition}
The Nyström approximation involves storing a submatrix of the whole kernel matrix. Thus, storage and computational cost are reduced to $O(nm)$ and $O(nm^2)$ respectively.
Nyström works by selecting $m<n$ points, callled representative points; it approximates K as
\begin{equation}
    \tilde {K}=K_{n,m} K_{m,m}^{-1}K_{n,m}^\intercal
\end{equation}

\subsubsection{Pivoted Cholesky decomposition}
Pivoted Cholesky approximate the Cholesky decomposition of a matrix. Since kernel matrices are positive definite, they can decomposed in terms of the Cholesky decomposition. Hence, we have that pivoted Cholesky can be used to approximate the full kernel matrix.

### 2024-03-27 thesis/input_files/paper1.tex
March 27th 2024

>>>>

\subsection{Kernel families}
Table \ref{tab:kernel types} contains popular kernel families in literature and applications.
\begin{table}
    \caption{Kernel types}
    \begin{tabular}{lll}
        \toprule
       Kernel function & Equation & Hyperparameters \\
       \midrule
       Linear &  $k(x_1,x_2)=x_1x_2$ &   \\
       Polynomial &  $k(x_1,x_2)=(x_1^\intercal x_2+c)^d$ &   c, d\\
       Gaussian RBF &  $k(x_1,x_2)=e^{-\frac{\|x_1-x_2|\|^2}{2\sigma^2}}$ &   $\sigma$\\
       Exponential RBF/Laplacian& $k(x_1,x_2)=e^{-\frac{|x_1-x_2|}{\gamma}}$ &   $\gamma$\\
       Hyperbolic/Sigmoid Kernel &  $k(x_1,x_2)=\tanh(\gamma x_1^\intercal x_2+r)$ &  $\gamma$, r \\
       Periodic &  $k(x_1,x_2)=e^{\frac{-2 sin^2\left(\frac{\pi}{p}|x_1-x_2| \right)}{l^2}} $ &   p, l\\
        \bottomrule
    \end{tabular}
    \label{tab:kernel types}
\end{table}

### 2024-03-27 thesis/input_files/point.tex
March 18th 2024
March 26th 2024
March 27th 2024

\section{Generalized additive model}

\section{K-nearest neighbors}

@@

@@

\section{Kernel methods}
\subsection{Kernel regression}
\subsection{Kernel support vector regression}

>>>>

\section{Generalized additive model}
%prophet meta
\section{K-nearest neighbors}

@@

Developed at the AT\&T Bell Laboratories by Vapnik et al. \cite{cortes1995support}, \cite{vapnik1997support}, support vector machines SVMs are one of the most popular techniques within the field of statistical learning.
\\
The goal of support vector regression is finding a function f(x) with at most $\epsilon$ deviation from the actual observed data $y_i$ $\forall i$ and as flat as possible. 
Put differently, we would like a model to keep error less than the $\epsilon$ threshold,  
%https://stats.stackexchange.com/questions/5945/understanding-svm-regression-objective-function-and-flatness
In standard SVR we have
\begin{equation}
    f(x)=\langle w,x \rangle +b \ \textrm{with} \ w \in X, b \in R
\end{equation}
Where $X$ denotes the space of the input patters $x_i$.
We can translate the flatness requirement into minimising the squared norm of w. Doing so we can formulate our problem as a convex optimisation problem.
\begin{equation}
    \begin{aligned}
        \min \quad& \frac{1}{2}\|w\|^2
        \\
        s.t. \quad& y_i-\langle w, x_i\rangle-b\leq \epsilon
        \\
        \quad& \langle w, x_i\rangle +b-y_i\leq \epsilon
    \end{aligned}
\end{equation}
Next, we need to introduce the slack variables $\xi$ and $\xi^*$, in order to handle the above formulate optimisation problem.
We obtain the following formulation
\begin{equation}
    \begin{aligned}
        \min \quad& \frac{1}{2}\|w\|^2+C\sum\limits_{i=1}^l(\xi_i+\xi_i^*)
        \\
        s.t. \quad& y_i-\langle w, x_i\rangle-b\leq \epsilon+\xi
        \\
        \quad& \langle w, x_i\rangle +b-y_i\leq \epsilon+\xi^*
        \\
        \quad& \xi_i\geq0
        \\
        \quad& \xi_i^*\geq0
    \end{aligned}
\end{equation}
The C constant trades off between $\epsilon$ deviation tolerance and flatness of the function f.
We seek to minimise the epsilon insensitive loss function, defined as follows
\begin{equation}
    \|\xi\|_\epsilon:=\begin{cases}
        0 \quad& \textrm{if} \ \|\xi\|\leq \epsilon
        \\
        \|\xi\|-\epsilon \quad& \textrm{if} \ \|\xi\|\leq \epsilon
    \end{cases}
\end{equation}
The model is depicted in figure \ref{fig:svm_simple}. The gry band is called the epsilon insensitive tube, only the points outside it are accounted by the loss function.
\begin{figure}
    \includegraphics[width=\textwidth]{images/svm_simple.png}    
    \caption{Support vector regression,\cite{learning_with_kernels}}
    \label{fig:svm_simple}
\end{figure}
Smola et al. \cite{smola2004tutorial} point out that considering the dual formulation makes our optimisation problem easier to solve. Doing so we have
\begin{equation}
    \begin{aligned}
        \max \quad& -\frac{1}{2}\sum\limits_{i,j=1}^l(\alpha_i-\alpha_i^*)(\alpha_j-\alpha_j^*)\langle x_i,x_j\rangle +\sum\limits_{i=1}^l(\alpha_i-\alpha_i^*)(y_i-\epsilon)
        \\
        s.t. \quad& \sum\limits_i=1^l(\alpha_i-\alpha_i^*)=0
        \\
        \quad& \alpha_i, \alpha_i^* \in [0, C]
    \end{aligned}
\end{equation}
Rearranging the gradient of the lagrangian with respect to w, we obtain the so called support vector expansion.
\begin{equation}\label{eq:support vector expansion}
    w=\sum\limits_{i=1}^l(\alpha_i-\alpha_i^*)x_i
\end{equation}
That means that w can be completely described as a linear combination of the training features $x_i$.
Notice, the complexity of the function representation is independent of the feature space dimensionaly but depends only on the number of support vectors; that is those point $i$ for which $(\alpha_i-\alpha_i^*)\neq 0$
Notice that, we do not needd to compute w explicitly in order to evaluate f(x).
Furthermore, equation \ref{eq:support vector expansion} implies
\begin{equation}
    f(x)=\sum\limits_{i=1}^l(\alpha_i-\alpha_i^*)\langle x_i, x\rangle +b
\end{equation}
Employ the Karush-Kuhn-Tucker conditions, b can be retrieved easily. Particularly, use the condition that the product between the constraints and the dual variable has to vanish. These implies that the lagrange multipliers $\alpha_i, \alpha_i^*$ may be nonzero only for the samples inside the epsilon insensitive tube. Consequently for any of these data points, the equality $b=y_i-\langle w, x_i-\epsilon\rangle$ holds.
To get an idea, see figure \ref{fig:svr1} for a how support vector regression handles a sinusoidal function with noise.
\\
\begin{figure}
    \includegraphics[width=\textwidth]{images/svr1.png}
    \caption{support vector regression}
    \label{fig:svr1}
\end{figure}

@@

\section{Kernel methods}
\subsection{Kernel regression}
\subsection{Kernel support vector regression}
Support vector regression can be kernelized by swapping the R2 euclidean dot product of data x with the dot product in the higher feature space F. Doing so, the optimisation problem can be restated as
\begin{equation}
    \begin{aligned}
        \max \quad& -\frac{1}{2}\sum\limits_{i,j=1}^l(\alpha_i-\alpha_i^*)(\alpha_i-\alpha_i^*)k(x_i, x_j)+\sum\limits_{i=1}^l(y_i-\epsilon)(\alpha_i-\alpha_i^*)
        \\
        s.t. \quad& \sum\limits_{i=1}^l(\alpha_i-\alpha_i^*)=0
        \\
        \quad& \alpha_i, \alpha_i^* \in [0,C]
    \end{aligned}
\end{equation}
Our regressor f is then given by
\begin{equation}
    \begin{aligned}
        \sum\limits_{i=1}^l(\alpha_i-\alpha_i^*)k(x_i, x)+b
    \end{aligned}
\end{equation}
Notice, in this setting w is no longer given explicitly and we seek for the flattest function in the feature space not the input space.
See figure \ref{fig:svr2} and figure \ref{fig:svr3} for two examples. In the former we have used a polynomial kernel while in the latter the standard radial basis function kernel.
\begin{figure}
    \includegraphics[width=\textwidth]{images/svr2.png}
    \caption{polynomial support vector regression}
    \label{fig:svr2}
\end{figure}

\begin{figure}
    \includegraphics[width=\textwidth]{images/svr3.png}
    \caption{rbf support vector regression}
    \label{fig:svr3}
\end{figure}
Comparing theese pictures with figure \ref{fig:svr1}, it can be concluded that introducing kernels allows support vector regression to handle non linearities in the data.
